SIRV-RL: Reinforcement Learning for Optimized Policy Control during Epidemiological Outbreaks in Emerging Market and Developing Economies

### 2024-02-17 paper_draft/main.tex
some more edits

In general, the policy $\pi$ is a mapping from states to a probability distribution over actions: $\pi: \altmathcal{S} \rightarrow p(\altmathcal{A}=\mathbf{a} \mid \altmathcal{S})$. If the MDP is episodic, i.e., the state is reset after each episode of length $T$, then the sequence of states, actions and rewards in an episode constitutes a trajectory or rollout of the policy. Every rollout of a policy accumulates rewards from the environment, resulting in the return $R=\sum_{t=0}^{t=T-1} \gamma^t r_{t+1}$. The goal of RL is to find an optimal policy, $\pi^*$, which achieves the maximum expected return from all states. To achieve this, reinforcement learning start with an initial arbitrary policy, i.e., a $Q$-table with no entries. $Q$-table is a mapping from states $s_t \in \mathscr{S}$ to a predefined set of actions to increase or decrease the stringency at time t, which are the actions $a_t \in \mathscr{A}$. Each entry of the $Q$-table $\left(Q_t\left(s_t, a_t\right)\right)$ associates an action in the finite sequence $\left(\mathscr{A}_j\right)_{j \in \mathbb{J}^{+}}$ to a state of the finite sequence $\left(\mathscr{S}_i\right)_{i \in \mathbb{I}^{+}}$~\cite{sutton2018reinforcement}.

@@

With this basis we define the reward function~\cite{Ohi2020}. In Deep Reinforcement Learning (DRL), positive rewards promote and negative rewards demote actions. The agent tries to generate such a policy/knowledge to avoid the discouraging situation by following the policy. By designing a proper reward function, it is possible to generate such an agent that may follow the human desired situation. While designing a reward function it is important to note that the rewards we set up truly indicate what we want accomplished. In particular, the reward signal is not the place to impart to the agent prior knowledge about how to achieve what we want it to do~\cite{sutton2018reinforcement}.
% TODO add: http://www.incompleteideas.net/book/ebook/node29.html

The reward function is parameterized to account for key factors influencing decision-making. To incentivize reduction of $R_e$ (effective reproductive number) the reward is $((R_e^{2}) \times -100) + 100$, this means when $R_e = 1$ there is no reward, but it exponentially decreases when $R_e > 1$ and increases if $R_e < 1$. Similarly, for the normalized GDP the reward function is $((\textrm{min\_max\_normalized\_GDP} ** 5)*300) - 80$ which tends to become negative below the average

If the proportion of the infected population were to rise above 0.003 the model is punished ($-5000$) and otherwise rewarded ($+20$). To reward not changing the stringencies frequently, we reward the absolute different between the previous stringency and the current stringency negatively ($|s(t) - s(t-1)| * -20$).

It should be realized there can be an infinite number of ways to design the reward function to be more human and upgrade the way a decision is taken given the situation \cite{AWSDeepRacer}. Therefore, this research act as a  framework for promoting the development of more efficient reward strategies for the same.

The agent observes the percentage of the population that is susceptible, infected, recovered, and the GDP which extrapolated from the stringency index and all the past moves played, i.e., all the past stringency indexes decided by the agent. The following values are fed into a simple network. Stable Baselines3 supports multiple inputs by using Dict Gym space. This can be done using MultiInputPolicy, which by default uses the CombinedExtractor features extractor to turn multiple inputs into a single vector, handled by the network arch network. For data that varies with time (stringency, normalized GDP, $R_e$) we use a simple LSTM architecture \cite{LSTM}. For other data like, the current proportion of the population that is susceptible, infected, and recovered we use a simple fully connected layer. The output from both these networks are concatenated and used by the reinforcement learning agent to train on.

@@

With a time-varying $\nu$ (vaccination rate) and the effect of lockdown, our model is able to account for the infected individuals and reduce the cost in comparison to all the previously formalized models for the data. This shows how interventions and changes in the way people behave in response of a epidemic \cite{Caldwell2021} play a major role in the way the epidemic unfolds.

@@

\section{Experiment Settings}
\subsection{Dataset}
The population-level epidemiological data can be obtained from the 'Our World In Data COVID-19' dataset: \url{https://ourworldindata.org/coronavirus} or more specifically: \url{https://github.com/owid/covid-19-data/blob/master/public/data/owid-covid-data.csv}~\cite{owidcoronavirus}. Data for the total cases, and recovered was acquired by scraping the Worldometers website~\cite{WorldometerCorona} used Internet Archive \cite{InternetArchive}. Quaterly GDP data can be obtained from the 'Organisation for Economic Co-operation and Development': \url{https://www.oecd-ilibrary.org/economics/data/main-economic-indicators/main-economic-indicators-complete-database_data-00052-en}~\cite{economic_indicators_data}.

>>>>

Given that at each timestep $t$ the agent has to choose an action $a_t$ so as to maximize the reward $r_{t+1}$ a policy is formulized by the agent. The policy $\pi$ is a mapping from states to a probability distribution over actions: $\pi: \altmathcal{S} \rightarrow p(\altmathcal{A}=\mathbf{a} \mid \altmathcal{S})$. Reinforcement learning methods specify how the agent changes its policy as a result of its experience. If the MDP is episodic, i.e., the state is reset after each episode of length $T$, then the sequence of states, actions and rewards in an episode constitutes a trajectory or rollout of the policy. Every rollout of a policy accumulates rewards from the environment, resulting in the return $R=\sum_{t=0}^{t=T-1} \gamma^t r_{t+1}$. The goal of RL is to find an optimal policy, $\pi^*$, which achieves the maximum expected return from all states. To achieve this, reinforcement learning start with an initial arbitrary policy, i.e., a $Q$-table with no entries. $Q$-table is a mapping from states $s_t \in \mathscr{S}$ to a predefined set of actions to increase or decrease the stringency at time t, which are the actions $a_t \in \mathscr{A}$. Each entry of the $Q$-table $\left(Q_t\left(s_t, a_t\right)\right)$ associates an action in the finite sequence $\left(\mathscr{A}_j\right)_{j \in \mathbb{J}^{+}}$ to a state of the finite sequence $\left(\mathscr{S}_i\right)_{i \in \mathbb{I}^{+}}$~\cite{sutton2018reinforcement}.

@@

In reinforcement learning, positive rewards promote and negative rewards demote actions. The agent tries to generate such a policy/knowledge to avoid the discouraging situation by following the policy. By designing a proper reward function, it is possible to generate such an agent that may follow the human desired situation. While designing a reward function it is important to note that the rewards we set up truly indicate what we want accomplished. In particular, the reward signal is not the place to impart to the agent prior knowledge about how to achieve what we want it to do~\cite{sutton2018reinforcement}. Taking inspiration from similar work~\cite{Ohi2020}, we define the reward function. % TODO add: http://www.incompleteideas.net/book/ebook/node29.html

The reward function is parameterized to account for key factors influencing decision-making. To incentivize reduction of $R_e$ (effective reproductive number) the reward is $((R_e^{2}) \times -100) + 100$, this means when $R_e = 1$ there is no reward, but it exponentially decreases when $R_e > 1$ and increases if $R_e < 1$. Similarly, for the normalized GDP the reward function is $((\textrm{min\_max\_normalized\_GDP}^5) \times 300) - 80$ which tends to become negative somewhere below the average of the $\textrm{min\_max\_normalized\_GDP}$ which is $0.75$. Since these are competing costs when added together for the real data they either cancel each other out for the duration of the pandemic and towards the end when both objectives are met they add up together.
% TODO: add image here
If the proportion of the infected population were to rise above 0.003 (peak in the actual data) the model is punished ($-2000$) and otherwise rewarded ($+20$). To reward not changing the stringencies frequently, we reward the absolute different between the previous stringency and the current stringency negatively ($|s(t) - s(t-1)| \times -20$).

It should be realized there can be an infinite number of ways to design the reward function to be more human and upgrade the way a decision is taken given the situation~\cite{AWSDeepRacer}. Therefore, this research act as a  framework for promoting the development of more efficient reward strategies for the same.

The agent observes the percentage of the population that is susceptible, infected, recovered, and time varying data like the GDP, and previous actions taken to change the stringency, and $R_e$. Since Stable Baselines3 can support multiple inputs(time-series data, single data points and images) by using Dict Gym space. We use the CombinedExtractor features extractor to turn multiple inputs into a single vector, handled by the network arch network. For data that varies with time (stringency, normalized GDP, $R_e$) we use a simple LSTM architecture~\cite{LSTM}. For other data like, the current proportion of the population that is susceptible, infected, and recovered we use a simple fully connected layer. The output from both these networks are concatenated and used by the reinforcement learning agent to train on.

@@

With a time-varying $\nu$ (vaccination rate) and the effect of lockdown, our model is able to account for the infected individuals and reduce the cost in comparison to all the previously formalized models for the data. This shows how interventions and changes in the way people behave in response of a epidemic~\cite{Caldwell2021} play a major role in the way the epidemic unfolds.

@@

\section{Experiment Settings}
\subsection{Dataset}
The population-level epidemiological data can be obtained from the 'Our World In Data COVID-19' dataset: \url{https://ourworldindata.org/coronavirus} or more specifically: \url{https://github.com/owid/covid-19-data/blob/master/public/data/owid-covid-data.csv}~\cite{owidcoronavirus}. Data for the total cases, and recovered was acquired by scraping the Worldometers website~\cite{WorldometerCorona} used Internet Archive~\cite{InternetArchive}. Quaterly GDP data can be obtained from the 'Organisation for Economic Co-operation and Development': \url{https://www.oecd-ilibrary.org/economics/data/main-economic-indicators/main-economic-indicators-complete-database_data-00052-en}~\cite{economic_indicators_data}.
